---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# AwesomePackage

<!-- badges: start -->
<!-- badges: end -->

The goal of AwesomePackage is to infer ancestry with some models and fit those models with some algorithms.

We use the classical PSD model for ancestor inference, which has been widely used, such as STRUCTURE (Pritchard et al. 2000, MCMC), FRAPPE (Tang et al. 2005, EM), ADMIXTURE (Alexander et al. 2009, SQP), fastSTRUCTURE (Raj et al. 2014, VI), TeraStructure (Gopalan et al. 2017, SVI). We illustrate the close relationship between the PSD model, the Poisson NMF model, the multinomial topic model and the LDA model, which can optimize the algorithm. We use Expectation-Maximization algorithm (EM), sequential quadratic programming algorithm (SQP), variational inference algorithm (VI) and stochastic variational inference algorithm (SVI) to fit the model, then illustrate the relationships and differences between these algorithms through simulation experiments and real data experiments.

## Installation

You can install the development version of AwesomePackage from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("JONATHONCHOW/AwesomePackage")
```

## Example

You can check out theories and examples at *Articles* in [AwesomePackage](https://jonathonchow.github.io/AwesomePackage/).

## Quick start

You can use the following code to see if AwesomePackage has been successfully installed.

```{r}
library(AwesomePackage)
hello_world()
```

You can refer to *Reference* in [AwesomePackage](https://jonathonchow.github.io/AwesomePackage/) for the use of functions, and then you can have fun with ancestry inference!
